---
title: "Phenotypic selection models"
author: "Francisco E. Fonturbel"
date: "5/October/2020"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Testing new approaches to estimate phenotypic selection in wild plant populations

This file contains the code to reproduce the statistical analyses presented in the paper.

```{r data, message=FALSE, warning=FALSE, include=FALSE}
##Libraries
library(mgcv)
library(corrplot)
library(stats)
library(ncf)
library(spdep)
library(mpmcorrelogram)
library(adegenet)
library(PopGenReport)
library(rsq)
library(ggplot2)
library(ggsci)
library(GGally)
library(mgcv.helper)

#Adding multiplot capabilities
source("multiplot.R")

##Some ggplot2 grooming
My_Theme = theme(
  axis.title.x = element_text(size = 20,face="bold"),
  axis.text.x = element_text(size = 18),
  axis.text.y = element_text(size = 18),
  axis.title.y = element_text(size = 20,face="bold"),
  legend.text = element_text(size = 10),
  legend.title = element_text(size = 10,face="bold"),
  plot.title = element_text(size = 16,face="bold"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.background = element_rect(fill = "white"))

##Loading data
data_sci<-read.table(file='data/dataSciRep.txt',header=T,sep='',dec='.')
#Loading microsatellite data (load QM69_microsats.gtx when asked)
microsats<-read.genetix(file='data/QM69_microsat.gtx')
data_rel<-gd.smouse(microsats)
```

## Correlation among traits

First, we will examine the correlation among traits measured (based on their standardized values):

```{r corr, echo=FALSE, message=FALSE, warning=FALSE}
##Correlation version 2.0
mm<-stats::cor(data_sci[,6:8], use="pairwise.complete.obs")
cortrait<-ggcorr(data = NULL, cor_matrix = cor(mm, use = "everything")) +
  My_Theme
        
multiplot(cortrait)
```

```{r corrs, message=FALSE, warning=FALSE, include=FALSE}
fruit.cor<-cor(data_sci$fruitS, data_sci$seedS, method="pearson")
sugar.cor<-cor(data_sci$fruitS, data_sci$sugarS, method="pearson")
seed.cor<-cor(data_sci$seedS, data_sci$sugarS, method="pearson")
```


As we can see here, those fruit traits present an important correlation, which makes sense from the biological point of view.

Fruit and seed size are **positively correlated** (i.e., larger fruits contain larger seeds), with a Pearson correlation coefficient of `r round(fruit.cor,3)`. On the contrary, fruit size and sugar content are **negatively correlated** (i.e., larger fruits contain less sugar concentration), with a Pearson correlation coefficient of `r round(sugar.cor,3)`. There is also a correlation between seed size and sugar content with a Pearson correlation coefficient of `r round(seed.cor,3)`, but maybe it is biologically meaningless...

### Trait relationships with fitness (w)

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=12}
plot.ws<-ggplot(data = data_sci, aes(x = seedS, y = w)) +
  geom_smooth(method = "glm", size = 1, span = 0.75, col = "red") +
  labs(x="Seed size", y="Relative fitness (w)") +
  theme_classic() +
  My_Theme
plot.wf<-ggplot(data = data_sci, aes(x = fruitS, y = w)) +
  geom_smooth(method = "glm", size = 1, span = 0.75, col = "blue") +
  labs(x="Fruit size", y="Relative fitness (w)") +
  theme_classic() +
  My_Theme
plot.wz<-ggplot(data = data_sci, aes(x = sugarS, y = w)) +
  geom_smooth(method = "glm", size = 1, span = 0.75, col = "chartreuse3") +
  labs(x="Sugar content", y="Relative fitness (w)") +
  theme_classic() +
  My_Theme

multiplot(plot.ws, plot.wf, plot.wz, cols = 1)
```

## Multidimensional Scaling

Now we will conduct a Multidimensional Scaling (MDS) Analysis (after Gower,1966) to summarize relatedness data in two dimensions.

```{r mds, echo=FALSE, message=FALSE, warning=FALSE}
##Multidimensional Scaling (MDS) or Principal Coordinates Analysis (Gower,1966)
mds <- cmdscale((data_rel), k = 2)
mds.plot<-ggplot(data = data_sci, aes(x = mds[,1], y = mds[,2])) +
  geom_point(col = "blueviolet", size = 3) +
  labs(x="MDS 1", y="MDS 2") +
  theme_classic() +
  My_Theme
mds.plot
```

This MDS vector will be used in further analyses to take the genetic component into account.

# Models
## GAM models

The first approach that we will try is based on Generalized Additive Models `gam` that include the elements of Lande & Arnold's equation plus two splines: one to take the space into account and the other to take the inter-individual relatedness into account.

```{r gam, message=FALSE, warning=FALSE, include=FALSE}
##GAM Models
model1<-gam(w ~ fruitS+seedS+sugarS,data=data_sci,family=gaussian)
model2<-gam(w ~ fruitS+seedS+sugarS+s(x,y,k=10),data=data_sci,family=gaussian)
model3<-gam(w ~ fruitS+seedS+sugarS+s(mds[,1],mds[,2],k=7),data=data_sci,family=gaussian)
model4<-gam(w ~ fruitS+seedS+sugarS+s(x,y,k=10)+s(mds[,1],mds[,2],k=8),data=data_sci,family=gaussian)

##BIC
#anova(model1,model2,model3,model4)
#BIC(model1,model2,model3,model4)

##Residuals of GAM
res1<-residuals(model1)
res2<-residuals(model2)
res3<-residuals(model3)
res4<-residuals(model4)

##Residual splines of GAM
cres1<-spline.correlog(x=data_sci$x,y=data_sci$y,z=res1, resamp=1000)
cres2<-spline.correlog(x=data_sci$x,y=data_sci$y,z=res2, resamp=1000)
cres3<-spline.correlog(x=data_sci$x,y=data_sci$y,z=res3, resamp=1000)
cres4<-spline.correlog(x=data_sci$x,y=data_sci$y,z=res4, resamp=1000)
```

**Classic model**

```{r model1, echo=FALSE, message=FALSE, warning=FALSE}
summary(model1)
```

**Spatial model**

```{r model2, echo=FALSE, message=FALSE, warning=FALSE}
summary(model2)
```

**Genetic model**

```{r model3, echo=FALSE, message=FALSE, warning=FALSE}
summary(model3)
```

**Spatial and genetic model**

```{r model4, echo=FALSE, message=FALSE, warning=FALSE}
summary(model4)
```

**Model comparison:**

```{r gam_perf, echo=FALSE, message=FALSE, warning=FALSE}
anova(model1,model2,model3,model4)
BIC(model1,model2,model3,model4)
```

## Selection gradients

Now we'll use the `termplot` function to plot selection gradients.

### Classic model

```{r term1, echo=FALSE, message=FALSE, warning=FALSE}
termplot(model1, rug = TRUE, se = TRUE, col.se = "gray", col.term = "black", lwd.term = 3, lwd.se = 2)
```

### Spatial model

```{r term2, echo=FALSE, message=FALSE, warning=FALSE}
termplot(model2, rug = TRUE, se = TRUE, col.se = "gray", col.term = "black", lwd.term = 3, lwd.se = 2)
```

### Genetic model

```{r term3, echo=FALSE, message=FALSE, warning=FALSE}
termplot(model3, rug = TRUE, se = TRUE, col.se = "gray", col.term = "black", lwd.term = 3, lwd.se = 2)
```

### Spatial + genetic model

```{r term4, echo=FALSE, message=FALSE, warning=FALSE}
termplot(model4, rug = TRUE, se = TRUE, col.se = "gray", col.term = "black", lwd.term = 3, lwd.se = 2)
```

**Note:** These plots can be found in the _termplot_ folder.

## GAM check

We will use the `gam.check` function to see how good (or bad) our models are bot overfitted (i.e., too many parameters for the data) and how residuals are behaving.

### Classic model

```{r check1, echo=FALSE, message=FALSE, warning=FALSE}
gam.check(model1, old.style = TRUE, k.rep = 1000)
```

### Spatial model

```{r check2, echo=FALSE, message=FALSE, warning=FALSE}
gam.check(model2, old.style = TRUE, k.rep = 1000)
```

### Genetic model

```{r check3, echo=FALSE, message=FALSE, warning=FALSE}
gam.check(model3, old.style = TRUE, k.rep = 1000)
```

### Spatial + genetic model

```{r check4, echo=FALSE, message=FALSE, warning=FALSE}
gam.check(model4, old.style = TRUE, k.rep = 1000)
```

## Assessing multicollinearity

To be sure that our results are not influenced by strong multicollinearity, we performed three tests: (1) estimating confidence intervals, (2) estimating the variance inflation factors (VIF), and (3) a `concurvity` analysis (except for the classic model).

**About concurvity:** Concurvity occurs when some smooth term in a model could be approximated by one or more of the other smooth terms in the model. This is often the case when a smooth of space is included in a model, along with smooths of other covariates that also vary more or less smoothly in space. Similarly it tends to be an issue in models including a smooth of time, along with smooths of other time varying covariates.

Concurvity can be viewed as a generalization of co-linearity, and causes similar problems of interpretation. It can also make estimates somewhat unstable (so that they become sensitive to apparently innocuous modelling details, for example).

This routine computes three related indices of concurvity, all bounded between 0 and 1, with 0 indicating no problem, and 1 indicating total lack of identifiability. The three indices are all based on the idea that a smooth term, f, in the model can be decomposed into a part, g, that lies entirely in the space of one or more other terms in the model, and a remainder part that is completely within the term's own space. If g makes up a large part of f then there is a concurvity problem. The indices used are all based on the square of ||g||/||f||, that is the ratio of the squared Euclidean norms of the vectors of f and g evaluated at the observed covariate values.

### Classic model

```{r mcol1, echo=FALSE, message=FALSE, warning=FALSE}
confint(model1)
vif.gam(model1)
```

### Spatial model

```{r mcol2, echo=FALSE, message=FALSE, warning=FALSE}
confint(model2)
vif.gam(model2)
concurvity(model2)
```

### Genetic model

```{r mcol3, echo=FALSE, message=FALSE, warning=FALSE}
confint(model3)
vif.gam(model3)
concurvity(model3)
```

### Spatial + genetic model

```{r mcol4, echo=FALSE, message=FALSE, warning=FALSE}
confint(model4)
vif.gam(model4)
concurvity(model4)
```

## Autocorrelation plots

```{r gam_plots, echo=FALSE, fig.width=8, message=FALSE, warning=FALSE}
###Respective plots of GAM
par(mfrow=c(2,2))
plot(cres1, main="(a) Standard model")
plot(cres2, main="(b) Spatial model")
plot(cres3, main="(c) Genetic model")
plot(cres4, main="(d) Spatial and genetic model")
par(mfrow=c(1,1))

#Export in 6x8 landscape mode

par(mfrow=c(1,2))
plot(correlog(data_sci$x,data_sci$y,data_sci$w,increment=2000))
plot(correlog(mds[,1],mds[,2],data_sci$w,increment=3))
par(mfrow=c(1,1))
```

### Mantel partial correlograms

Testing significance using Mantel partial correlograms:

```{r mpm, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
xy<-cbind(data_sci$x,data_sci$y)
mdist<-dist(xy)

#Making distance matrices upon residuals
mres1<-dist(res1)
mres2<-dist(res2)
mres3<-dist(res3)
mres4<-dist(res4)

mpm_c1<-mpmcorrelogram(mres1,mdist,method="pearson",permutations=10000)
mpm_c2<-mpmcorrelogram(mres2,mdist,method="pearson",permutations=10000)
mpm_c3<-mpmcorrelogram(mres3,mdist,method="pearson",permutations=10000)
mpm_c4<-mpmcorrelogram(mres4,mdist,method="pearson",permutations=10000)
```

```{r mpm_plots, echo=FALSE, message=FALSE, warning=FALSE}
par(mfrow=c(2,2))
plot(mpm_c1, xlab="Distance classes",main="(a) Standard model")
plot(mpm_c2, xlab="Distance classes",main="(b) Spatial model")
plot(mpm_c3, xlab="Distance classes",main="(c) Genetic model")
plot(mpm_c4, xlab="Distance classes",main="(d) Spatial and genetic model")
par(mfrow=c(1,1))
```

Oh boy! the genetic model deals **better** with autocorrelation than the spatial model! that's very interesting!

## Session

```{r session, echo=T}
sessionInfo()
```